---
title: "Tenth World Congress of Melanoma in Conjuction with 17th EADO Congress"
subtitle: "April 15-17, 2021"
author: 
date: "`r format(Sys.Date(), '%d %B %Y')`"
fontfamily: mathpazo
geometry: 'margin=1in'
fontsize: '12pt'
toc: false
linestretch: 1.5
output: 
  officedown::rdocx_document:
    number_section: false
    base_format: bookdown::word_document2
    reference_docx: 'Final-Submitted/Abstract.docx'
  bookdown::html_document2:
    number_sections: false
    theme: yeti
  bookdown::pdf_document2:
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  comment = NA, 
  warning = FALSE,
  message = FALSE
)
library(data.table)
library(gtsummary)

gtsummary::theme_gtsummary_compact()
gtsummary::theme_gtsummary_journal('jama')
```

```{r data}
data <- readRDS("../../../Documents/Papers/Analysis/data/extended.rds")
```

```{r}
make_table_summary <- function(data) {
  data[, 
       .(AgeDiag, BreslowThickness, BreslowTstage, YearCat)
  ] %>% tbl_summary(by = YearCat, label = list(
    AgeDiag ~ "Age at diagnosis",
    BreslowThickness ~ "Tumour thickness",
    BreslowTstage ~ "T category"
  ), missing_text = 'Unspecified') %>% 
    modify_header(update = all_stat_cols() ~ fcase(
      knitr::is_html_output(), "**{level}**<br>N = {n}", 
      knitr::is_latex_output(), "**{level}**\nN = {n}",
      default = "**{level}**\nN = {n}"
    ))
}
```

**Title:** Long-term trends in melanoma tumour thickness in Norway

**Authors:** Raju Rimal^1^, Trude E Robsahm^2^, Adele Green^3,4^, Reza Ghiasvand^5^, Corina S Rueegg^5^, Assia Bassarova^6^, Petter Gjersvik^7^, Elisabete Weiderpass^8^, Odd O Aalen^1^, Bjørn Møller^9^, Marit B Veierød^1^

**Affiliations:**

^1^  Oslo Centre for Biostatistics and Epidemiology, Department of Biostatistics, Institute of Basic Medical Sciences, University of Oslo, Oslo, Norway  

^2^  Department of Research, Cancer Registry of Norway, Oslo, Norway 

^3^  Department of Population Health, QIMR Berghofer Medical Research Institute, Brisbane, Australia 

^4^  Cancer Research UK Manchester Institute, University of Manchester, Manchester, United Kingdom 

^5^  Oslo Centre for Biostatistics and Epidemiology, Oslo University Hospital, Oslo, Norway 

^6^  Department of Pathology, Oslo University Hospital -- Ullevål, Oslo, Norway 

^7^  Institute of Clinical Medicine, University of Oslo, Oslo, Norway 

^8^  International Agency for Research on Cancer, Lyon, France 

^9^  Department of Registration, Cancer Registry of Norway, Oslo, Norway


**Categories:** Melanoma

**Topics:** Epidemiology

**Preference:** Poster Presentation

# Background

Norway has the second-highest mortality rate of cutaneous melanoma worldwide and ranks fifth in incidence. Tumour (Breslow) thickness at diagnosis is the primary determinant of the T category in the tumour, nodes, metastasis staging system, and the most important prognostic factor for survival after localized melanoma. This ongoing study investigates long-term trends in tumour thickness, and the corresponding T categories, overall and in important subgroups, in a nationwide case series over a 40-year time period.

# Methods

```{r}
nobs <- nrow(data)
```

The population-based Cancer Registry of Norway (CRN) provided all first primary invasive melanoma cases for 1980-2019. Tumour thickness was available from the Norwegian Melanoma Registry (within the CRN) for all cases diagnosed in 2008-2019 and was manually extracted from the paper notifications archived in the CRN for the cases diagnosed in 1980-2007. Covariates include sex, age, residential geographical region, anatomic site, histopathological subtype, clinical stage, and ulceration.

Descriptive summaries are presented as frequencies (numbers, %) and medians with interquartile ranges (IQR).

# Results

```{r}
as <- data[, .(table(AnatomicSite, Gender) %>% prop.table(margin = "Gender"))] %>% 
  .[, N := scales::percent(N, accuracy = 0.1)]
mt <- data[, .(table(MelanomaType) %>% prop.table())] %>% 
  .[, N := scales::percent(N, accuracy = 0.1)]
cs <- data[, .(table(ClinicalStage) %>% prop.table())] %>% 
  .[, N := scales::percent(N, accuracy = 0.1)]
```

In total, `r format(nobs, big.mark = ",")` morphologically verified first primary invasive melanoma cases were diagnosed in Norway in 1980-2019. In both men and women, median age at diagnosis increased from 1980-2000 to 2008-2019 (Table \@ref(tab:table1)). Trunk was the most common anatomic site in men (`r as[Gender == "Male" & AnatomicSite == "Trunk", N]`), and trunk (`r as[Gender == "Female" & AnatomicSite == "Trunk", N]`) and lower extremities (`r as[Gender == "Female" & AnatomicSite == "Lower extremities", N]`) the most common sites in women. The dominant subtype was superficial spreading melanoma (`r mt[1, N]`), followed by nodular melanoma (`r mt[2, N]`). In total, `r cs[1, N]` of the cases were diagnosed in a local stage, whereas `r cs[2, N]` had regional metastases and `r cs[3, N]` had distant metastases.

```{r}
bt <- data[, lapply(list(
  Median = function(x) median(x, na.rm = TRUE),
  Q1 = function(x) quantile(x, 0.25, na.rm = TRUE),
  Q2 = function(x) quantile(x, 0.75, na.rm = TRUE)
), function(fn) fn(BreslowThickness)), by = .(YearCat, Gender)] %>% 
  .[, label := paste0(Median, " mm ", "(", paste(Q1, Q2, sep = "--"), ")")] %>% 
  .[YearCat %in% c('1980-1999', '2008-2019')]
```

Women were diagnosed at a thinner stage than men. In men, median (IQR) tumour thickness decreased from `r bt[Gender=="Male" & YearCat == "1980-1999", label]` in 1980-1999 to `r bt[Gender=="Male" & YearCat == "2008-2019", label]` in 2008-2019, and in women from `r bt[Gender=="Female" & YearCat == "1980-1999", label]` to `r bt[Gender=="Female" & YearCat == "2008-2019", label]` (Table \@ref(tab:table1)).

Tumour thickness was missing in the pathology reports for more than 25% of the cases until 1990. Reporting of ulceration started in 2000, but with a large proportion of missing values. After the Norwegian Melanoma Registry was established in 2008, the proportions of missing ulceration decreased dramatically.

```{r table1}
gtsummary::theme_gtsummary_compact()
gtsummary::theme_gtsummary_journal('jama')
tbl_male <- make_table_summary(data[Gender == "Male"])
tbl_female <- make_table_summary(data[Gender == "Female"])
tbl <- gtsummary::tbl_merge(
  list(Male = tbl_male, Female = tbl_female), 
  tab_spanner = c("**Male**", "**Female**")) %>%
  modify_footnote(update = everything() ~ NA) %>%
  modify_footnote(update = label ~ "Median (IQR); n (%)") %>% 
  modify_caption("Age and tumour thickness in Norwegian melanoma cases1, 1980-2019.")
tbl$table_body <- tbl$table_body %>% 
  dplyr::filter(ifelse(variable == "BreslowTstage", label != "Unspecified", TRUE))
if (knitr::is_html_output()) {
  tbl %>% bold_labels() %>% 
    as_kable_extra(escape = FALSE) %>% 
    kableExtra::kable_styling(font_size = 12)
} else if (knitr::is_latex_output()) {
  tbl %>% 
    as_kable_extra(format = "latex", booktabs = TRUE) %>% 
    kableExtra::kable_paper("striped", full_width = TRUE) %>%
    kableExtra::column_spec(1, bold = TRUE)
} else {
  tbl %>% bold_labels() %>% 
    as_flex_table() %>%
    flextable::theme_vanilla() %>% 
    flextable::bold(part = "header") %>%
    flextable::set_table_properties(layout = "autofit") %>% 
    flextable::fontsize(part = "all", size = 8) %>% 
    flextable::set_caption(
      autonum = officer::run_autonum(),
      caption = "Age and tumour thickness in Norwegian melanoma cases1, 1980-2019."
    )
}
```

Analysis of incidence rates in relation to tumour thickness is in the process, and will be presented at the conference.

# Conclusions

This unique time series of national melanoma tumour thickness data will identify trends in tumour thickness, overall and in subgroups of the population, as well as identify potential effects of changing exposure patterns and earlier detection.
